Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-02 tagless/tagless-final.tex
Some simple BibTeX work

\hyphenation{meta-language meta-languages Meta-OCaml}
\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage[T1]{fontenc}
\usepackage[override]{cmtt}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1}
\usepackage{mdwlist}

@@

If our OCaml-like metalanguage were untyped, the code above would be acceptable.
The |L e| line exhibits interpretive overhead:
|eval0| traverses the function body~|e| every time (the result of
evaluating) |L e| is applied. Staging can be used to remove this
interpretive overhead \cite[\S1.1--2]{WalidICFP02}.

@@

Besides immediate evaluation, we can compile our object language
into OCaml code using MetaOCaml's staging facilities. MetaOCaml
represents future-stage expressions of type |t| at the
present stage as values of type |('c,t) code| where |'c| is the
environment classifier \cite{WalidPOPL03,calcagno-ml-like}. Code values are created
by a \emph{bracket} form |.<e>.|, which specifies that the expression |e| is to be
evaluated at a future stage. The \emph{escape} |.~e| must occur
within a bracket and specifies that the expression |e| must be evaluated
at the current stage; its result, which must be a code value, is
spliced into the code being built by the enclosing bracket. The \emph{run} form |.!e| evaluates
the future-stage code value~|e| by compiling and linking it at run-time.
The bracket, escape, and run are akin to
quasi-quotation, unquotation, and |eval| of Lisp.

@@

Our |Symantics| class includes only separate lifting methods |bool| and
|int|, not a parametrically polymorphic lifting method, for good reason:
When compiling to a first-order target language such as machine code,
booleans, integers, and functions may well be represented differently.
Thus polymorphic lift cannot be compiled without intensional type
analysis.  To avoid the need for polymorphic lift, we turn to
a technique due to Asai \ccshan{cite}: build a dynamic term
alongside every static term.

@@

The object language generally inherits the evaluation strategy from
the metalanguage. Evaluating an application |e1 e2| in OCaml requires
the evaluation of |e2| before performing the application. Therefore,
when evaluating the encoding of the object language application 
|app e1 e2|, |e2| will always have to be evaluated. One may wonder if it is
possible to represent a call-by-name (CBN) object language in a
call-by-value (CBV) metalanguage. The answer, given already by 
\cite{ReynoldsPlotkin} is continuation-passing style (CPS), which they
specifically introduced 
to make the evaluation strategy of a definitional interpreter
independent of the strategy of the metalanguage. We demonstrate the
same independence in the typed setting \oleg{mention the quote from
Reynolds that typed setting would be nice?}\jacques{I think so, but from
which paper? The one ReynoldsPlotkin paper is typed}. Specifically, we
demonstrate CBN evaluation for our object language in OCaml, which is
call-by-value. We do so by building a CBN CPS interpreter for the object
language.

The CBN CPS interpretation was introduced by Plotkin \cite{PlotkinCBN}. As in
the more familiar CBV CPS, the interpretation of a value is a function
accepting a continuation argument |k| and yielding the answer
returned by that continuation.
\begin{code}
let int (x:int) = fun k -> k x
let add e1 e2 = 
  fun k -> e1 (fun v1 -> e2 (fun v2 -> k (v1+v2)))
\end{code}
In both cases, the returned value has type 
|(int -> 'w)->'w| where |'w| is the (polymorphic) answer type.

@@

We can easily change our CBN CPS interpreter into a CBV CPS
interpreter, by replacing only one definition:
\begin{code}
module RCV = struct
  include RCN
  let lam f = {ko = 
    fun k -> k (fun e -> e.ko 
         (fun v -> f {ko = fun k -> k v}))}
end;;
\end{code}
Now when we apply an abstraction to an argument |e|, we always force
the evaluation of that argument before proceeding. This approach is in
line with \cite{ReynoldsXXX} (albeit in a typed setting). The interpreter RCV
is useful for achieving CBV evaluation of the object language
no matter whether the metalanguage is CBV or CBN.

@@

\section{Self-interpretation}\cite{selfinterp}

@@

oleg{Mention the RR interpreter: for any term E, (RR E) is equivalent to E.
RR is not an identity: it is an interpreter that encapsulates another
interpreter. The RR interpreter can be made self if we treat RR as a
special form and interpret it always with itself (similar to let and
hole below).}

@@

\bibliographystyle{mcbride}
\bibliography{tagless}
\end{document}

>>>>

\hyphenation{meta-language meta-languages Meta-OCaml meta-cir-cu-lar}
\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage[T1]{fontenc}
\usepackage[override]{cmtt}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1}
\usepackage{mdwlist}

@@

If our OCaml-like metalanguage were untyped, the code above would be acceptable.
The |L e| line exhibits interpretive overhead:
|eval0| traverses the function body~|e| every time (the result of
evaluating) |L e| is applied. Staging can be used to remove this
interpretive overhead \citep[\S1.1--2]{WalidICFP02}.

@@

Besides immediate evaluation, we can compile our object language
into OCaml code using MetaOCaml's staging facilities. MetaOCaml
represents future-stage expressions of type |t| at the
present stage as values of type |('c,t) code| where |'c| is the
environment classifier \citep{WalidPOPL03,calcagno-ml-like}. Code values are created
by a \emph{bracket} form |.<e>.|, which specifies that the expression |e| is to be
evaluated at a future stage. The \emph{escape} |.~e| must occur
within a bracket and specifies that the expression |e| must be evaluated
at the current stage; its result, which must be a code value, is
spliced into the code being built by the enclosing bracket. The \emph{run} form |.!e| evaluates
the future-stage code value~|e| by compiling and linking it at run-time.
The bracket, escape, and run are akin to
quasi-quotation, unquotation, and |eval| of Lisp.

@@

Our |Symantics| class includes only separate lifting methods |bool| and
|int|, not a parametrically polymorphic lifting method, for good reason:
When compiling to a first-order target language such as machine code,
booleans, integers, and functions may well be represented differently.
Thus polymorphic lift cannot be compiled without intensional type
analysis.  To avoid the need for polymorphic lift, we turn to
\citearound{'s technique}\citet[see also \citealp{sumii-hybrid}]{asai-binding-time}:
build a dynamic term
alongside every static term.

@@

The object language generally inherits the evaluation strategy from
the metalanguage. Evaluating an application |e1 e2| in OCaml requires
the evaluation of |e2| before performing the application. Therefore,
when evaluating the encoding of the object language application 
|app e1 e2|, |e2| will always have to be evaluated. One may wonder if it is
possible to represent a call-by-name (CBN) object language in a
call-by-value (CBV) metalanguage. The answer, given already by 
\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}, is continuation-passing style (CPS), which they
specifically introduced 
to make the evaluation strategy of a definitional interpreter
independent of the strategy of the metalanguage. We demonstrate the
same independence in the typed setting \oleg{mention the quote from
Reynolds that typed setting would be nice?}\jacques{I think so, but from
which paper? The one ReynoldsPlotkin paper is typed}\ccshan{I think Oleg
meant ``It should also be possible to define languages with a highly
 refined syntactic type structure. Ideally, such a treatment should be
 metacircular, in
 the sense that the type structure used in the defined language should be adequate for the
 defining language'' \citep{reynolds-definitional}, but perhaps Jacques
 meant another paper by ``ReynoldsPlotkin''?}. Specifically, we
demonstrate CBN evaluation for our object language in OCaml, which is
call-by-value. We do so by building a CBN CPS interpreter for the object
language.

The CBN CPS interpretation was introduced by \citet{PlotkinCBN}. As in
the more familiar CBV CPS, the interpretation of a value is a function
accepting a continuation argument |k| and yielding the answer
returned by that continuation.
\begin{code}
let int (x:int) = fun k -> k x
let add e1 e2 = 
  fun k -> e1 (fun v1 -> e2 (fun v2 -> k (v1+v2)))
\end{code}
In both cases, the returned value has type 
|(int -> 'w)->'w| where |'w| is the (polymorphic) answer type.

@@

We can easily change our CBN CPS interpreter into a CBV CPS
interpreter, by replacing only one definition:
\begin{code}
module RCV = struct
  include RCN
  let lam f = {ko = 
    fun k -> k (fun e -> e.ko 
         (fun v -> f {ko = fun k -> k v}))}
end;;
\end{code}
Now when we apply an abstraction to an argument |e|, we always force
the evaluation of that argument before proceeding. This approach is in
line with \citet{reynolds-relation} (albeit in a typed setting). The interpreter RCV
is useful for achieving CBV evaluation of the object language
no matter whether the metalanguage is CBV or CBN.

@@

\section{Self-interpretation}\label{selfinterp}

@@

\oleg{Mention the RR interpreter: for any term E, (RR E) is equivalent to E.
RR is not an identity: it is an interpreter that encapsulates another
interpreter. The RR interpreter can be made self if we treat RR as a
special form and interpret it always with itself (similar to let and
hole below).}

@@

\bibliographystyle{mcbride}
\bibsep=0pt
\bibliography{tagless}
\end{document}
